\documentclass[]{article}
\usepackage{url}
\usepackage{booktabs}

%opening
\title{Reviewer Responses to TrustCom15 Submission}
\author{Andrew Bolster}

\begin{document}

\maketitle

\begin{abstract}
We regret to inform you that your paper could not be accepted for inclusion in the technical program of
TrustCom2015 conference. However, based on the reviewers' reports, your paper is considered as a
good quality paper. WE ARE PLEASED TO INFORM YOU that your paper has been invited for RATSP 2015
(The 2015 IEEE International Symposium on Recent Advances of Trust, Security and Privacy in
Computing and Communications https://research.comnet.aalto.fi/Trustcom2015/ratsp2015/), which
will be held in conjunction with TrustCom 2015. The RATSP 2015 papers will be included in the
TrustCom 2015 proceedings to be published by IEEE CS press.
\end{abstract}

\section{Score Summary}

\begin{tabular}{|l|c|c|c||c}
	\toprule
	\textbf{Area}        & \textbf{R1} & \textbf{R2} & \textbf{R3} \\ \midrule
	Significance (10\%)  &      3      &      4      &      3      \\
	Novelty (20\% )      &      2      &      4      &      2      \\
	Structure (10\%)     &      4      &      4      &      3      \\
	Justification (10\%) &      3      &      5      &      3      \\
	General (50 \%)      &      3      &      4      &      3      \\ \midrule
	Overall              &      1      &      2      &     -1      \\
	Confidence           &      4      &      3      &      4      \\ \bottomrule& 
\end{tabular} 

\section{Review 1}

This paper simulates underwater autonomous network environment, showing a detailed comparative analysis for MTMF, OTMF and Hermes program, as well as a metric weighting analysis for MTMF. 

As a whole, \textbf{the structure of the paper is rather rigorous}, and a lot of experiments have been done.

However, though this paper aims to apply the Trust Management Frameworks to Underwater Autonomous Networks, in the paper a large number of experimental data was not collected under the real underwater environment of the autonomous network, but instead a simulated environment, which makes the data lack of reliability. \textbf{In addition, this paper is not innovative enough. }

Regardless of MTMF，OTMF or Hermes, they are existing technology. \textbf{The main contribution of this paper is just a study on the comparative operation and performance of TMFs in marine acoustic networks.}

Thus, I recommend that this paper should be weak accepted.

\section{Review 2}

Many thanks to the authors for their efforts. In this paper, the authors propose a framework to address the challenges for trust management in the under-water environment. \textbf{The paper is well organised and the presentation is good. }

This paper is easy to follow. In addition, the experiments are enough to demonstrate the advantages by using a multi-metric approach in the underwater environment. 

There is only one comment for this paper: I feel the \textbf{contributions of this paper are not well presented in the introductio}n part so that I have to search in the whole context. 

\textbf{Could you please improve the introduction part a bit?} Anyway, I recommend to accept this paper.

\section{Review 3}

This paper aims at providing a study on the comparative operation and performance of trust management frameworks in marine acoustic networks, and a review of metric suitability for TMFs in marine environments. 

However, from this paper's references, it is evident that most of the simulation results were repeating the work in [6] which was designed for mobile ad hoc networks.

Section II discussed various trust management frameworks in different networks. 
However, in subsection C. Single Metric Trust Frameworks, this paper only discussed one framework which was from reference [7]; at the same time, for subsection D. Multi-Metric Trust Frameworks, it only discussed one framework which was from reference [6]. 

\textbf{Without the comparison with other frameworks, how did the paper justify that the selected metrics were more suitable for marine acoustic networks.}

The organization of this paper is not good as well. For example, in V. SIMULATION RESULTS AND DISCUSSION, it is \textbf{not clear how to identify the malicious behaviors in nodes}.

\end{document}
